---
title: "Fitting structural mean models"
author: "Tom Palmer"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 4
    code_download: true
    code_folding: show
    anchor_sections: true
---

```{r include=FALSE}
require("Hmisc")
```
`r Hmisc::hidingTOC(levels = 4)`

# Introduction

This is the code from the appendix of Clarke, et al., Estimating Structural Mean Models with
Multiple Instrumental Variables Using the Generalised Method of Moments. Statistical Science, 2015, 30, 1, 96-177. https://dx.doi.org/10.1214/14-STS503

# Example data

Using the data from example 2 from the `ivtools::ivglm()` helpfile.

```{r}
set.seed(9)
n <- 1000
psi0 <- 0.5
psi1 <- 0.2
Z <- rbinom(n, 1, 0.5)
X <- rbinom(n, 1, 0.7*Z + 0.2*(1 - Z)) 
m0 <- plogis(1 + 0.8*X - 0.39*Z)
Y <- rbinom(n, 1, plogis(psi0*X + log(m0/(1 - m0)))) 
dat <- data.frame(Z, X, Y)
```

# Fitting Structural Mean Models

* Setup: we need the gmm and ivtools packages
```{r setup, message=FALSE}
library(gmm)
library(ivtools)
library(logbin)
library(boot)
library(ivreg)
library(msm)
```

## Additive structural mean model: estimation of the causal risk difference

* Formula syntax
```{r}
asmm <- gmm(dat[, "Y"] ~ dat[, "X"], x = dat[, c("Z")], vcov = "iid")

# model summary
summary(asmm)

# estimates with 95% CIs
cbind(coef(asmm), confint(asmm)$test)

# over-identification test
# specTest(asmm)
```

* Alternative function syntax
```{r}
asmmMoments <- function(theta, x){
  # extract variables from x
  Y <- x[,"Y"]
  X <- x[,"X"]
  Z1 <- x[,"Z"]
  # moments
  m1 <- (Y - theta[1] - theta[2]*X)
  m2 <- (Y - theta[1] - theta[2]*X)*Z1
  return(cbind(m1, m2))
}

asmm2 <- gmm(asmmMoments, x = dat, t0 = c(0, 0), vcov = "iid")

# estimates
summary(asmm2)
cbind(coef(asmm2), confint(asmm2)$test)

# over-identification test
# specTest(asmm2)
```

### Comparison fits

#### Two-stage residual inclusion
* TSRI estimated using GMM with additive error
```{r}
tsriIdentAddMoments <- function(theta, x){
  # extract variables from x
  Y <- x[,"Y"]
  X <- x[,"X"]
  Z1 <- x[,"Z"]
  # generate first stage residuals
  stage1 <- lm(X ~ Z1)
  res <- residuals(stage1)
  # moments
  a1 <- (X - theta[1] - Z1*theta[2])
  a2 <- (X - theta[1] - Z1*theta[2])*Z1
  m1 <- (Y - (theta[3] + X*theta[4] + theta[5]*(X - theta[1] - Z1*theta[2])))
  m2 <- (Y - (theta[3] + X*theta[4] + theta[5]*(X - theta[1] - Z1*theta[2])))*X
  m3 <- (Y - (theta[3] + X*theta[4] + theta[5]*(X - theta[1] - Z1*theta[2])))*res
  return(cbind(a1, a2, m1, m2, m3))
}

tsrigmmident <- gmm(tsriIdentAddMoments, x = dat, t0 = rep(0,5), vcov = "iid")

# estimates
summary(tsrigmmident)

# causal risk difference
cbind(coef(tsrigmmident), confint(tsrigmmident)$test)[3:5,]
```

#### G-estimation
```{r}
fitZ.L <- glm(Z ~ 1, data = dat)
fitY.LZX <- glm(Y ~ X + Z + X*Z, family = "binomial", data = dat)
fitIdentGest <-
  ivglm(
    estmethod = "g",
    X = "X",
    fitZ.L = fitZ.L,
    fitY.LZX = fitY.LZX,
    data = dat,
    link = "identity",
    Y = "Y"
  )

# model summary
summary(fitIdentGest)

# estimates and CIs
cbind(fitIdentGest$est, confint(fitIdentGest))

# plot estimating function
estf1 <- estfun(fitIdentGest)
plot(estf1)
```

## Multiplication structural mean model: estimation of the causal risk ratio

```{r}
msmmMoments <- function(theta, x){
  # extract variables from x
  Y <- x[,"Y"]
  X <- x[,"X"]
  Z1 <- x[,"Z"]
  # moments
  m1 <- (Y*exp(-1*X*theta[2]) - theta[1])
  m2 <- (Y*exp(-1*X*theta[2]) - theta[1])*Z1
  return(cbind(m1, m2))
}

msmm <- gmm(msmmMoments, x = dat, t0 = c(0, 0), vcov = "iid")

# estimates
summary(msmm)

# causal risk ratio
exp(cbind(coef(msmm), confint(msmm)$test)[2,])

# E[Y(0)]
cbind(coef(msmm), confint(msmm)$test)[1,]

# over-identification test
# specTest(msmm)
```

* Alternative syntax

```{r}
msmmAltMoments <- function(theta, x){
  # extract variables from x
  Y <- x[,"Y"]
  X <- x[,"X"]
  Z1 <- x[,"Z"]
  # moments
  m1 <- (Y*exp(-theta[1] - X*theta[2]) - 1)
  m2 <- (Y*exp(-theta[1] - X*theta[2]) - 1)*Z1
  return(cbind(m1, m2))
}

msmm2 <- gmm(msmmAltMoments, x = dat, t0 = c(0, 0), vcov = "iid")

# exponentiate estimates
exp(cbind(coef(msmm2), confint(msmm2)$test))

# over-identification test
# specTest(msmm2)
```

### Comparison fits

#### Two-stage residual inclusion
```{r, cache=TRUE, message=FALSE, warning=FALSE}
stage1 <- lm(X ~ Z, data = dat)
dat$res <- residuals(stage1)
stage2 <- logbin(Y ~ X + res, data = dat)

# model summary
summary(stage2)

# estimates and (perhaps slightly incorrect) CIs
exp(cbind(coef(stage2), confint(stage2)))

# checking CIs using bootstrap SEs

stat <- function(data, indices) {
  stage1 <- lm(X ~ Z, data = dat, subset = indices)
  dat$res[indices] <- residuals(stage1)
  stage2 <- logbin(Y ~ X + res, data = dat, subset = indices)
  return(coef(stage2))
}

boot(data = dat, statistic = stat, R = 100)
```

* TSRI estimated using GMM with multiplicative error
```{r}
tsriLogMultMoments <- function(theta, x){
  # extract variables from x
  Y <- x[,"Y"]
  X <- x[,"X"]
  Z1 <- x[,"Z"]
  # generate first stage residuals
  stage1 <- lm(X ~ Z1)
  res <- residuals(stage1)
  # moments
  a1 <- (X - theta[1] - Z1*theta[2])
  a2 <- (X - theta[1] - Z1*theta[2])*Z1
  m1 <- (Y*exp(-theta[3] - X*theta[4] - theta[5]*(X - theta[1] - Z1*theta[2])) - 1)
  m2 <- (Y*exp(-theta[3] - X*theta[4] - theta[5]*(X - theta[1] - Z1*theta[2])) - 1)*X
  m3 <- (Y*exp(-theta[3] - X*theta[4] - theta[5]*(X - theta[1] - Z1*theta[2])) - 1)*res
  return(cbind(a1, a2, m1, m2, m3))
}

tsrigmmlog1 <- gmm(tsriLogMultMoments, x = dat, t0 = rep(0,5), vcov = "iid")

# estimates
summary(tsrigmmlog1)

# causal risk ratio
exp(cbind(coef(tsrigmmlog1), confint(tsrigmmlog1)$test)[3:5,])
```

* TSRI estimated using GMM with additive error
```{r}
tsriLogAddMoments <- function(theta, x){
  # extract variables from x
  Y <- x[,"Y"]
  X <- x[,"X"]
  Z1 <- x[,"Z"]
  # generate first stage residuals
  stage1 <- lm(X ~ Z1)
  res <- residuals(stage1)
  # moments
  a1 <- (X - theta[1] - Z1*theta[2])
  a2 <- (X - theta[1] - Z1*theta[2])*Z1
  m1 <- (Y - exp(theta[3] + X*theta[4] + theta[5]*(X - theta[1] - Z1*theta[2])))
  m2 <- (Y - exp(theta[3] + X*theta[4] + theta[5]*(X - theta[1] - Z1*theta[2])))*X
  m3 <- (Y - exp(theta[3] + X*theta[4] + theta[5]*(X - theta[1] - Z1*theta[2])))*res
  return(cbind(a1, a2, m1, m2, m3))
}

tsrigmmlog2 <- gmm(tsriLogAddMoments, x = dat, t0 = rep(0,5), vcov = "iid")

# estimates
summary(tsrigmmlog2)

# causal risk ratio
exp(cbind(coef(tsrigmmlog2), confint(tsrigmmlog2)$test)[3:5,])
```

#### G-estimation
```{r}
fitZ.L <- glm(Z ~ 1, data = dat)
fitY.LZX <- glm(Y ~ X + Z + X*Z, family = "binomial", data = dat)
summary(fitY.LZX)

fitLogGest <-
  ivglm(
    estmethod = "g",
    X = "X",
    fitZ.L = fitZ.L,
    fitY.LZX = fitY.LZX,
    data = dat,
    link = "log",
    Y = "Y"
  )

# model summary
summary(fitLogGest)

# estimates and CIs
exp(cbind(fitLogGest$est, confint(fitLogGest)))

# plot the estimating function
estf <- estfun(fitLogGest)
plot(estf)
```

#### TSLS using transformed variables

* Equations 24 and 25 of Clarke et al., i.e. outcome Y(1 - X) with exposure YX

```{r}
dat$yxm1 <- dat$Y * (1 - dat$X)
dat$yx <- dat$Y * dat$X
dat$mz <- 1 - dat$Z

# first stage
stage1 <- glm(yx ~ Z, data = dat, family = binomial)
summary(stage1)
dat$hatyx <- fitted.values(stage1)

# tsls fit
tslsmsmm <- ivreg(yxm1 ~ yx | Z, data = dat)
summary(tslsmsmm)

# causal risk ratio estimate
beta <- coef(tslsmsmm)
-1 / beta[2]
logcrr <- log(-1 / beta[2])

# delta-method SE
estvar <- vcov(tslsmsmm)
logcrrse <- deltamethod(~ log(-1 / x2), beta, estvar)
logcrrse

# crr with 95% CI
res <- unname(c(-1/beta[2], exp(logcrr - 1.96*logcrrse), exp(logcrr + 1.96*logcrrse)))
res
```

## Logistic structural mean model: estimation of the causal risk ratio

```{r}
# association model
am <- glm(Y ~ X + Z + X*Z, as.data.frame(dat), family = binomial)
print(summary(am))
amfit <- coef(am)
xblog <- qlogis(fitted.values(am))

# causal model with incorrect SEs
cmMoments <- function(theta, x){
  # extract variables from x
  X <- x[,"X"]
  Z1 <- x[,"Z"]
  # moments
  c1 <- (plogis(xblog - theta[2]*X) - theta[1])
  c2 <- (plogis(xblog - theta[2]*X) - theta[1])*Z1
  return(cbind(c1, c2))
}

cm <- gmm(cmMoments, x = dat, t0 = c(0, 0), vcov = "iid")
cmfit <- coef(cm)

lsmmMoments <- function(theta, x){
  # extract variables from x
  Y <- x[,"Y"]
  X <- x[,"X"]
  Z1 <- x[,"Z"]
  XZ1 <- X*Z1
  # association model moments
  xb <- theta[1] + theta[2]*X + theta[3]*Z1 + theta[4]*XZ1
  a1 <- (Y - plogis(xb))
  a2 <- (Y - plogis(xb))*X
  a3 <- (Y - plogis(xb))*Z1
  a4 <- (Y - plogis(xb))*XZ1
  # causal model moments
  c1 <- (plogis(xb - theta[6]*X) - theta[5])
  c2 <- (plogis(xb - theta[6]*X) - theta[5])*Z1
  return(cbind(a1, a2, a3, a4, c1, c2))
}

lsmm <- gmm(lsmmMoments, x = dat, t0 = c(amfit, cmfit), vcov = "iid")

# model summary
print(summary(lsmm))

# E[Y(0)]
print(cbind(coef(lsmm), confint(lsmm)$test)[5])

# exponentiate other estimates
print(exp(cbind(coef(lsmm), confint(lsmm)$test)[-5,]))

# over-identification test
# print(specTest(lsmm))
```

### Comparison fits

#### Two-stage predictor substitution
```{r}
fitX.LZ <- glm(X ~ Z, family = "binomial", data = dat)
fitY.LX <- glm(Y ~ X, family = "binomial", data = dat)
fitLogisticTSPS <-
  ivglm(
    estmethod = "ts",
    fitX.LZ = fitX.LZ,
    fitY.LX = fitY.LX,
    data = dat,
    ctrl = FALSE
  )

# model summary
summary(fitLogisticTSPS)

# estimates and CIs
exp(cbind(fitLogisticTSPS$est, confint(fitLogisticTSPS)))
```

#### Two-stage residual inclusion
```{r}
fitLogisticTSRI <-
  ivglm(
    estmethod = "ts",
    fitX.LZ = fitX.LZ,
    fitY.LX = fitY.LX,
    data = dat,
    ctrl = TRUE
  )

# model summary
summary(fitLogisticTSRI)

# estimates and CIs
exp(cbind(fitLogisticTSRI$est, confint(fitLogisticTSRI)))
```

#### G-estimation
```{r}
fitZ.L <- glm(Z ~ 1, data = dat)
fitY.LZX <- glm(Y ~ X + Z + X*Z, family = "binomial", data = dat)
fitLogisticGest <-
  ivglm(
    estmethod = "g",
    X = "X",
    fitZ.L = fitZ.L,
    fitY.LZX = fitY.LZX,
    data = dat,
    link = "logit"
  )

# model summary
summary(fitLogisticGest)

# estimates and CIs
exp(cbind(fitLogisticGest$est, confint(fitLogisticGest)))

# plot estimating function
estf3 <- estfun(fitLogisticGest)
plot(estf3)
```
